Modelling BCG vaccination in the UK

### 2019-08-06 rmds/04-tb-epi-england.Rmd
refresh TB epi figures
tb epi page break
tb epi fig positioning
tb epi fig labelling updated

In this chapter I explore the epidemiology of TB in England using routine data-sets with a particular focus on the impact of missing data, the mechanisms underlying that missing data, seasonal trends, the role of age, UK birth status and BCG status. I also estimate incidence rates, stratified by UK birth status and age, which I use to identify trends in TB incidence over time. Finally I report TB outcomes in England using case rates, again stratified by BCG status and UK birth status. These data are then used throughout this thesis to explore the impact of BCG vaccination on TB outcomes (Chapter \@ref(beneficial-bcg-out)), to estimate the direct impact of the 2005 change in BCG vaccination policy (Chapter \@ref(direct-eff)), to parameterise a dynamic TB transmission model (Chapter \@ref(model-development)) and to fit the same model to data (Chapter \@ref(model-fitting)).

@@

Data on all notifications (114,820 notifications) from the ETS system from 2000 to 2015 were obtained from PHE via an application to the TB monitoring team. Data fields included: notification date, age, PHE centre, occupation, ethnic group, UK birth status, years since entry to the UK, date of symptom onset, date of presentation, date of diagnosis, date of treatment start, date of treatment end, date of death, Pulmonary TB status, culture status, sputum smear status, drug resistance, BCG vaccination status, year of vaccination, outcome at 12 months, overall outcome, and cause of death. Notifications were assessed for identifiability and the data release was conditional on the raw data not being shared further. Invalid entries were replaced with missing values unless otherwise noted, with character variables stored as factors using their most common entry as the baseline. Notifications from Scotland, Northern Ireland and Wales were dropped from the data-set. Several variables were created, or modified, for use in further analysis, Table \@ref(tab:created-var) summarises these variables. The code used for data cleaning is available as an R package^[Data cleaning code: https://www.samabbott.co.uk/tbinenglanddataclean/reference/clean_munge_ets_2016.html].

@@

As the ETS is aggregated across England, from a variety of sources, some level of missing data are inevitable. This takes two forms: under-reporting of notified cases, of which there is some evidence in the literature,[@Pillaye2003] and data missing for a notified case. The former is particularly problematic as apart from using comparative studies the characteristics of those that are not notified is unknown. For variables that are missing data within the data-set it is possible to calculate the proportion of missing data (Figure \@ref(fig:plot-missing-struct), Table \@ref(tab:missing-var-tabs)) but care must be taken to account for nested variables such as date of death and year of BCG vaccination. This can be done by assuming that the nested variables takes the value of the top level variable when it is known that the variable is not truly missing. An example of this is using overall outcome for date of death when notifications are known to have not died.  Doing this shows high completeness for common demographic variables such as sex, age, ethnic group and UK birth status. More problematically, BCG status and year of BCG status have a high percentage missing, even after accounting for the introduction of national collection of these variables in 2008. Socio-economic status (as national quintiles) was not collected until 2010 but after this point is highly complete.  Comparing pre 2009 and post 2008 in Table \@ref(tab:missing-var-tabs) (and by inspecting Figure \@ref(fig:plot-missing-struct)) there are also issues of changing completeness over time,[@PHE2016; @PHE2017] if this is not accounted for than it may lead to spurious trends. Figure \@ref(fig:plot-missing-struct) also indicates that there are multiple groups of variables that share a common pattern of missing data.

@@

For nested variables with rare outcomes assuming the top level variable value can mask the underlying amount of missing data. An alternative approach is to filter the data for the top level variable required for the nested variable to be defined and to then compute the proportion of these notifications that are missing data for the outcome of interest. For the date of starting treatment this approach leads to an estimate of `r missing_stats[["date_treat"]]` being missing, which is more complete than previously estimated. For cases that are known to have completed treatment `r missing_stats[["date_treat_end"]]` are missing a date for the end of treatment. In notifications that are known to have died, `r missing_stats[["date_death"]]` were missing the date of death and `r missing_stats[["cause_death"]]` were missing the cause of death. In any analysis were these variables are used the missing data for these variables will need to be carefully adjusted for. In particular, if cause of death is used it must be clearly stated that it is highly missing and results based on this variable should be properly caveated.

@@

```{r bcgvacc-miss}
kable(model_miss_results[["bcgvacc"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that BCG status is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, 
                latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r bcgvaccyr-miss}
kable(model_miss_results[["bvcvaccyr"]], format = "latex",
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that year of BCG vaccination is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm")
```

@@

```{r dateofdeath-miss}
kable(model_miss_results[["dateofdeath"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that there is some evidence that date of death is missing at random for ethnic group, with weaker evidence for all other variables.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r tomdeathrealat-miss}
kable(model_miss_results[["tomdeathrealat"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that cause of death is missing at random for ethnic group and UK birth status, with little evidence for any other variables", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r symptonset-miss}
kable(model_miss_results[["symptonset"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of symptom onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of symptom onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that date of symptom onset is missing not at random for the variables for all variables considered, except for sex.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "HOLD_position")
```

@@

```{r datediag-miss}
kable(model_miss_results[["datediag"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of diagnosis onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).", 
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of diagnosis onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that date of diagnosis is missing at random for the variables for all variables considered, except for sex.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

For date of starting treatment there is little evidence that missing data is associated with any variable considered, except for year of notification (Table \@ref(tab:starttreatdate-miss)). Variable completeness improved year on year, with a 96% drop in missing data in 2015 compared to 2010.

```{r starttreatdate-miss}
kable(model_miss_results[["starttreatdate"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).", 
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. There is little evidence that the missing data for the date of starting treatment is associated with any variable considered, except for year of notification.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>%   
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r txenddate-miss}
kable(model_miss_results[["txenddate"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis weas used. Odds ratios shown are adjusted for all explanatory variables. There is little evidence that the missing data for the date of starting treatment is associated with any variable considered, except for year of notification.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

Routine observational data-sets are subject to numerous potential biases, such as selection bias, recall bias, measurement bias, and unmeasured confounding.[@Benchimol2016a] Additionally as the data has not been collected with a specific analysis in mind there maybe issues with the specificity of variables. The ETS system is likely to suffer from all of the above biases to some extent, which must be accounted for as far as possbile, and explicitly stated at every level of analysis. The most important consideration is that the ETS system is unlikely to be representative of the general population as it contains only notified TB cases that occurred in England during the study period, research questions must therefore be either limited to active TB patients, or when extended to the general population the differing population demographics must be accounted for. If this is not done then any results may be due to selection bias. Additionally, multiple variables may suffer from misclassification bias, including BCG status which can be assessed via vaccination record, the presence of a scar, or case recall: this may lead to spurious associations.[@Fewell2007] Validation studies would be required to account for this, which is beyond the scope of this thesis.

@@

For analyses that aim to reproduce temporal trends in TB incidence, such as dynamic modelling studies, it is important to understand which variables represent the most accurate date of contact with the health system and more generally on what scale date variables can be considered reliable. In the ETS extract used in this thesis there are several date variables that encode useful information including: the date of notification, the date of symptom onset, the date of diagnosis, the date of starting treatment, the date of completing treatment, and the date of death. In the following section I explore these variables using counts and proportions aggregated to then nearest year, month and day.

@@

An alternative measure is to use the date of symptom onset, this represents the closest approximation to the date when a case became infectious. Unfortunately there are multiple issues with this measure, the first of which being is that it is only `r symptonset_complete` complete across the data extract. Additionally, completeness changes with time, with  `r complete_sym_year$Complete[1]` in 2000,  `r complete_sym_year$Complete[9]` in 2008, and `r complete_sym_year$Complete[15]` in 2014 this could lead to spurious trends in the number of cases. Perhaps most importantly the date of symptom onset is highly susceptible to recall bias with the majority of cases becoming symptomatic on the first of each month (Figure \@ref(fig:plot-detection-sympton) d.)), with some evidence that a greater number of cases occur in January than would be expected (Figure \@ref(fig:plot-detection-sympton) c.)). Another possible measure of the number of cases is the date of diagnosis, this should be a more reliable variable than the date of symptom onset, as it does not rely on the recall of the case. However it is only `r datediag_complete` complete across the dataset, with strong evidence of increasing completeness going from  `r complete_diag_year$Complete[1]` in 2000, `r complete_diag_year$Complete[9]` in 2008, and `r complete_diag_year$Complete[15]` in 2014. This trend would be hard to properly account for in any analysis and therefore this variable should not be used as a primary measure.

@@

The date of starting treatment should be a more reliable contact date as it records an official contact with the health system. Indeed it was `r complete_treatstart_year$Complete[1]` complete in 2000 which increased year on year to `r complete_treatstart_year$Complete[16]` complete in 2015. This increasing completeness may lead to a temporal bias if not properly adjusted for when evaluating the date of starting treatment over time. As for the data of notification there is some evidence of a seasonal trend for date of starting treatment (Figure \@ref(fig:plot-detection-starttreatdate) c.)), with a peak of cases starting treatment in May, June and July. However, this seasonal trend is difficult to identify when cases starting treatment are visualised by month over time (Figure \@ref(fig:plot-detection-starttreatdate) b.)). Unlike the date of symptom onset there is little evidence of recall bias by month, or by day (Figure \@ref(fig:plot-detection-starttreatdate) c.) and d.)).

@@

@@

Age-standardised rates were calculated using the `r 'epiR'` package for R,[@Stevenson2016] using the average age distribution of England, and Wales from 2000-2015 as the standard population to allow comparison between years. Those aged 90+ were excluded as demographic data for this population was unreliable. The code used to calculate incidence rates is available online as an R package (`tbinenglanddataclean`; see Chapter \@ref(introduction)).

@@

Case rates were calculated as follows and confidence estimates were estimated using the `prop.test` function from the `stats` package:

>>>>

In this chapter I explore the epidemiology of TB in England using routine datasets with a particular focus on the impact of missing data, the mechanisms underlying that missing data, seasonal trends, the role of age, UK birth status and BCG status. I also estimate incidence rates, stratified by UK birth status and age, which I use to identify trends in TB incidence over time. Finally I report TB outcomes in England using case rates, again stratified by BCG status and UK birth status. These data are then used throughout this thesis to explore the impact of BCG vaccination on TB outcomes (Chapter \@ref(beneficial-bcg-out)), to estimate the direct impact of the 2005 change in BCG vaccination policy (Chapter \@ref(direct-eff)), to parameterise a dynamic TB transmission model (Chapter \@ref(model-development)) and to fit the same model to data (Chapter \@ref(model-fitting)).

@@

Data on all notifications (114,820 notifications) from the ETS system from 2000 to 2015 were obtained from PHE via an application to the TB monitoring team. Data fields included: notification date, age, PHE centre, occupation, ethnic group, UK birth status, years since entry to the UK, date of symptom onset, date of presentation, date of diagnosis, date of treatment start, date of treatment end, date of death, Pulmonary TB status, culture status, sputum smear status, drug resistance, BCG vaccination status, year of vaccination, outcome at 12 months, overall outcome, and cause of death. Notifications were assessed for identifiability and the data release was conditional on the raw data not being shared further. Invalid entries were replaced with missing values unless otherwise noted, with character variables stored as factors using their most common entry as the baseline. Notifications from Scotland, Northern Ireland and Wales were dropped from the dataset. Several variables were created, or modified, for use in further analysis, Table \@ref(tab:created-var) summarises these variables. The code used for data cleaning is available as an R package^[Data cleaning code: https://www.samabbott.co.uk/tbinenglanddataclean/reference/clean_munge_ets_2016.html].

\newpage

@@

As the ETS is aggregated across England, from a variety of sources, some level of missing data are inevitable. This takes two forms: under-reporting of notified cases, of which there is some evidence in the literature,[@Pillaye2003] and data missing for a notified case. The former is particularly problematic as apart from using comparative studies the characteristics of those that are not notified is unknown. For variables that are missing data within the dataset it is possible to calculate the proportion of missing data (Figure \@ref(fig:plot-missing-struct), Table \@ref(tab:missing-var-tabs)) but care must be taken to account for nested variables such as date of death and year of BCG vaccination. This can be done by assuming that the nested variables takes the value of the top level variable when it is known that the variable is not truly missing. An example of this is using overall outcome for date of death when notifications are known to have not died.  Doing this shows high completeness for common demographic variables such as sex, age, ethnic group and UK birth status. More problematically, BCG status and year of BCG status have a high percentage missing, even after accounting for the introduction of national collection of these variables in 2008. Socio-economic status (as national quintiles) was not collected until 2010 but after this point is highly complete.  Comparing pre 2009 and post 2008 in Table \@ref(tab:missing-var-tabs) (and by inspecting Figure \@ref(fig:plot-missing-struct)) there are also issues of changing completeness over time,[@PHE2016; @PHE2017] if this is not accounted for than it may lead to spurious trends. Figure \@ref(fig:plot-missing-struct) also indicates that there are multiple groups of variables that share a common pattern of missing data.

@@

For nested variables with rare outcomes assuming the top level variable value can mask the underlying amount of missing data. An alternative approach is to filter the data for the top level variable required for the nested variable to be defined and to then compute the proportion of these notifications that are missing data for the outcome of interest. For the date of starting treatment this approach leads to an estimate of `r missing_stats[["date_treat"]]` being missing, which is more complete than previously estimated. For cases that are known to have completed treatment `r missing_stats[["date_treat_end"]]` are missing a date for the end of treatment. In notifications that are known to have died, `r missing_stats[["date_death"]]` were missing the date of death and `r missing_stats[["cause_death"]]` were missing the cause of death. In any analysis where these variables are used the missing data for these variables will need to be carefully adjusted for. In particular, if cause of death is used it must be clearly stated that it is highly missing and results based on this variable should be properly caveated.

@@

```{r bcgvacc-miss}
kable(model_miss_results[["bcgvacc"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that BCG status is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, 
                latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r bcgvaccyr-miss}
kable(model_miss_results[["bvcvaccyr"]], format = "latex",
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for year of BCG vaccination as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that year of BCG vaccination is missing at random for the variables considered.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm")
```

@@

```{r dateofdeath-miss}
kable(model_miss_results[["dateofdeath"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that there is some evidence that date of death is missing at random for ethnic group, with weaker evidence for all other variables.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

\newpage

```{r tomdeathrealat-miss}
kable(model_miss_results[["tomdeathrealat"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for cause of death as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that cause of death is missing at random for ethnic group and UK birth status, with little evidence for any other variables", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r symptonset-miss}
kable(model_miss_results[["symptonset"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of symptom onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of symptom onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that date of symptom onset is missing not at random for the variables for all variables considered, except for sex.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "HOLD_position")
```

@@

```{r datediag-miss}
kable(model_miss_results[["datediag"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of diagnosis onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).", 
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of diagnosis onset as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. The model indicates that date of diagnosis is missing at random for the variables for all variables considered, except for sex.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

For date of starting treatment there is little evidence that missing data is associated with any variable considered, except for year of notification (Table \@ref(tab:starttreatdate-miss)). Variable completeness improved year-on-year, with a 96% drop in missing data in 2015 compared to 2010.

```{r starttreatdate-miss}
kable(model_miss_results[["starttreatdate"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).", 
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. There is little evidence that the missing data for the date of starting treatment is associated with any variable considered, except for year of notification.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>%   
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

```{r txenddate-miss}
kable(model_miss_results[["txenddate"]],
      caption.short = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles).",
      caption = "Results from a logistic regression model with data completeness (Complete/Missing) for date of starting treatment as an outcome, adjusted for: year, sex, age (grouped as 0-14 year olds, 15-65 year olds and 65+), ethnic group, UK birth status and socio-economic status (national quintiles). For socio-economic group 1 indicates the most deprived quintile. Notifications from 2010 onwards were included as socio-economic status was not collected before this. Complete case analysis was used. Odds ratios shown are adjusted for all explanatory variables. There is little evidence that the missing data for the date of starting treatment is associated with any variable considered, except for year of notification.", booktabs = TRUE, longtable = TRUE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(c(1), width = "1.5cm") %>% 
  column_spec(c(4), width = "2cm") %>% 
  column_spec(c(6, 7), width = "1.5cm") %>% 
  kable_styling(latex_options = "hold_position")
```

@@

Routine observational datasets are subject to numerous potential biases, such as selection bias, recall bias, measurement bias, and unmeasured confounding.[@Benchimol2016a] Additionally, as the data has not been collected with a specific analysis in mind there maybe issues with the specificity of variables. The ETS system is likely to suffer from all of the above biases to some extent, which must be accounted for as far as possbile, and explicitly stated at every level of analysis. The most important consideration is that the ETS system is unlikely to be representative of the general population as it contains only notified TB cases that occurred in England during the study period, research questions must therefore be either limited to active TB patients, or when extended to the general population the differing population demographics must be accounted for. If this is not done then any results may be due to selection bias. Additionally, multiple variables may suffer from misclassification bias, including BCG status which can be assessed via vaccination record, the presence of a scar, or case recall: this may lead to spurious associations.[@Fewell2007] Validation studies would be required to account for this, which is beyond the scope of this thesis.

@@

For analyses that aim to reproduce temporal trends in TB incidence, such as dynamic modelling studies, it is important to understand which variables represent the most accurate date of contact with the health system and more generally on what scale date variables can be considered reliable. In the ETS extract used in this thesis there are several date variables that encode useful information including: the date of notification, the date of symptom onset, the date of diagnosis, the date of starting treatment, the date of completing treatment, and the date of death. In the following section I explore these variables using counts and proportions aggregated to the nearest year, month and day.

@@

An alternative measure is to use the date of symptom onset, this represents the closest approximation to the date when a case became infectious. Unfortunately there are multiple issues with this measure, the first of which being is that it is only `r symptonset_complete` complete across the data extract. Additionally, completeness changes with time, with  `r complete_sym_year$Complete[1]` complete in 2000,  `r complete_sym_year$Complete[9]` complete in 2008, and `r complete_sym_year$Complete[15]` complete in 2014. This could lead to spurious trends in the number of cases. Perhaps most importantly the date of symptom onset is highly susceptible to recall bias with the majority of cases becoming symptomatic on the first of each month (Figure \@ref(fig:plot-detection-sympton) d.)), with some evidence that a greater number of cases occur in January than would be expected (Figure \@ref(fig:plot-detection-sympton) c.)). Another possible measure of the number of cases is the date of diagnosis, this should be a more reliable variable than the date of symptom onset, as it does not rely on the recall of the case. However it is only `r datediag_complete` complete across the dataset, with strong evidence of increasing completeness going from  `r complete_diag_year$Complete[1]` complete in 2000, to `r complete_diag_year$Complete[15]` complete in 2014. This trend would be hard to properly account for in any analysis and therefore this variable should not be used as a primary measure.

@@

The date of starting treatment should be a more reliable contact date as it records an official contact with the health system. Indeed it was `r complete_treatstart_year$Complete[1]` complete in 2000 which increased year-on-year to `r complete_treatstart_year$Complete[16]` complete in 2015. This increasing completeness may lead to a temporal bias if not properly adjusted for when evaluating the date of starting treatment over time. As for the data of notification there is some evidence of a seasonal trend for date of starting treatment (Figure \@ref(fig:plot-detection-starttreatdate) c.)), with a peak of cases starting treatment in May, June and July. However, this seasonal trend is difficult to identify when cases starting treatment are visualised by month over time (Figure \@ref(fig:plot-detection-starttreatdate) b.)). Unlike the date of symptom onset there is little evidence of recall bias by month, or by day (Figure \@ref(fig:plot-detection-starttreatdate) c.) and d.)).

@@

\newpage

@@

Age-standardised rates were calculated using the `r 'epiR'` package for R,[@Stevenson2016] using the average age distribution of England from 2000-2015 as the standard population to allow comparison between years. Those aged 90+ were excluded as demographic data for this population were unreliable. The code used to calculate incidence rates is available online as an R package (`tbinenglanddataclean`^[`tbinenglanddataclean`: https://github.com/seabbs/tbinenglanddataclean]; see Chapter \@ref(introduction)).

@@

Case rates were calculated as follows and confidence estimates were estimated using the `prop.test` function from the `stats` R package:
